\RequirePackage[l2tabu,orthodox]{nag}
%
\documentclass[oneside,a4paper,oldfontcommands]{memoir}

%\newsubfloat{figure}

\input{settings}
\input{misc/acronyms.tex}
\makeglossaries

% \headstyles{ell}
\chapterstyle{ell}

% Use sans-serif font
\renewcommand{\kufrontfont}{\sffamily}

\project{\mdseries\LARGE Master's Thesis}
\author{Truls Asheim -- {\ttfamily truls@asheim.dk}}
%\title{A Common Intermediate Representation of Synchronous Message ExchangeNetworks}
%\title{This is a title}
\title{A Domain Specific Language for Synchronous Message Exchange Networks}
%\subtitle{ }
\date{May 2018}
\supervisor{Supervisors: Brian Vinter and Kenneth Skovhede}

\begin{document}

\frontmatter

\begin{titlingpage}
\maketitle
\end{titlingpage}

\newpage
\pagestyle{plain}
\begin{abstract}
  Synchronous Message Exchange (SME) is a CSP-derived model for hardware designs
  implementing globally synchronous message passing. SME implementations
  currently exist for several general-purpose languages, some of which, are
  translatable to VHDL for subsequent implementation on hardware. A common SME
  language could reduce the duplication and feature disparity present in these
  independent implementations. This thesis introduces a domain-specific language
  for implementing SME designs. It is usable both as a primary implementation
  language for SME models and as an intermediate target for general-purpose
  languages. We describe the language, its implementation and its
  features. Furthermore, we explain the specific requirements for a language
  within this domain. Finally, we evaluate the language through a number of
  simple, but realistic, hardware designs by showing how they may be implemented
  and tested.
\end{abstract}
\newpage


\newpage
\tableofcontents

\printglossary[type=\acronymtype,title=Abbrevations]

% \chapter{Preface}
% This thesis is submitted in fulfillment of the degree {\it Master of Computer
%   Science} for Truls Asheim.

\mainmatter

\input{chapters/introduction}
\input{chapters/sme}
\input{chapters/hdls}
%\input{chapters/analysis}
\input{chapters/smeil}
\input{chapters/cosim}
\input{chapters/codegen}
\input{chapters/implementation}
\input{chapters/evaluation}
\input{chapters/discussion}
\input{chapters/conclusion}

%\microtypesetup{protrusion=true}
\printbibliography{}


\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% TeX-command-extra-options: "-enable-write18"
%%% End:
