\chapter{\libsme{} design and implementation}

In this section, we present the combined implementation of \libsme{} and
elaborate on implementation details.

\section{Methods of interaction}
\label{sec:using}
SMEIL programs are run using the {\ttfamily libsme} library either through interaction
with the C API of the library or by using the provided command line utility.

\begin{description}
\item[Direct code generation.]
  % A SMEIL program constitutes a complete
  % description by itself provided that only size constrained types are
  % used.
  A SMEIL program that contains only size-constrained types provide all the
  information that is needed to generate a hardware description. Therefore, VHDL
  code can be generated directly from a SMEIL program without the intermediate
  simulation step. Some advantages are lost when using this mode, as no test
  bench is created and the generated VHDL code must be manually modified and
  connected to a clock source for driving the simulation before it can be tested
  using a VHDL simulator.
\item[Pure SMEIL simulation.] This mode only applies to SMEIL networks which
  contain their own data generation process (see \Cref{sec:7seg} and
  \Cref{sec:md5} for examples of such networks). SMEIL used like this is not
  very useful as it can only produce an output through {\ttfamily trace}
  statements (\Cref{sec:langref}).
\item[Co-simulation of SMEIL.] The most common intended usage scenario for SMEIL
  is to use it together with an SME library for a general purpose language. This
  allows the generation of VHDL code, an associated test bench, and a trace
  file. The full details of the co-simulation interface of SMEIL were previously
  given in \Cref{sec:cosim}.
\end{description}

\label{sec:libsmeimpl}
\begin{figure}%[tb]
  \centering
  \resizebox{\linewidth}{!}{
    \begin{tikzpicture}[font=\small,
      rep style/.style={rectangle,draw=black,text width=1.5cm,minimum
        size=1.5cm,align=center},
      proc style/.style={ellipse,draw=black,align=center,text
        width=1.5cm,minimum size=1.5cm,align=center},
      file style/.style={ellipse,draw=none,align=center,text
        width=1.5cm,minimum size=0.5cm,align=center}
      ]
      \node[proc style] (parser) {Parser};
      \node[proc style, right=1cm of parser] (import) {Import Resolver};
      \draw[-{Latex[scale=1.6]}] (parser) edge [bend right=30] (import);
      \draw[-{Latex[scale=1.6]},dashed] (import) edge [bend right=30] (parser);

      \node[proc style, right=1cm of import] (tyc) {Type Checker};
      \draw[-{Latex[scale=1.6]}] (import) edge (tyc);
      \node[proc style, below=3.1cm of tyc] (an) {Analyzer};

      \draw[-{Latex[scale=1.6]}] (tyc) edge [bend left=20] (an);
      \draw[-{Latex[scale=1.6]},dashed] (tyc) edge [bend right=20] (an);

      \node[proc style, below right=1cm and -0.5cm of import] (recon) {Retyper/\\Recon-\\structor};
      \node[proc style, below=0.5cm of parser] (sim) {Simulator};
      \node[proc style, below=1cm of sim] (codegen) {Code Generator};
      \draw[-{Latex[scale=1.6]}] (an) edge (sim);

      \draw[-{Latex[scale=1.6]}] (sim) edge (codegen);
      \draw[-{Latex[scale=1.6]},dashed] (sim) edge (recon);
      \draw[-{Latex[scale=1.6]},dashed] (recon) edge (tyc);
      \draw[-{Latex[scale=1.6]},dashed] (an) edge [bend left=20] (codegen);
      \draw[-{Latex[scale=1.6]}] (an) edge [bend right=10] (codegen);

      \node[rep style, left=2cm of parser, text width=2cm] (cmdl)  {{\ttfamily  smec} (Command line)};

      \node[rep style, below=0.5cm of cmdl, text width=2cm] (cosim)  {Co-simulator\\(PySME, ...)};
      \draw[-{Latex[scale=1.6]},dashed] (cosim) edge [bend right=10] node[above] {API} (sim);
      \draw[-{Latex[scale=1.6]},dashed] (sim) edge [bend right=10] (cosim);
      \draw[-{Latex[scale=1.6]}] (cosim) edge  (parser);

      \draw[-{Latex[scale=1.6]}] (cmdl) edge node[below=0.3cm,rotate=26]
      {{\ttfamily \tiny file.sme}} (parser);

      \node[file style, below=1cm of cosim] (trace) {Trace file};
      \node[file style, below=0.1cm of trace, text width=0.8cm] (vhdl) {VHDL};

      \draw[-{Latex[scale=1.6]}] (sim) edge (trace);
      \draw[-{Latex[scale=1.6]}] (codegen) edge (vhdl);

      \begin{scope}[on background layer]
        \node[draw=scopeborder,fill=scopebg,inner sep=10pt,rounded corners=10pt,anchor=north west,fit=(parser)(an)(codegen),label={below}:libsme] (libsme) {};
        \node[draw=scopeborder,fill=scopebg,inner sep=10pt,rounded corners=10pt,anchor=north west,fit=(cmdl)(cosim),label={above}:Interactors] (inter) {};
        \node[draw=scopeborder,fill=scopebg,inner sep=5pt,rounded corners=10pt,anchor=north west,fit=(trace)(vhdl),label={below}:Outputs] (inter) {};
      \end{scope}
    \end{tikzpicture}
  }
  \caption{Overviews of interactions with and data flow within \libsme{}. The
    dashed lines denotes paths which are followed conditionally depending on
    which mode \libsme{} is executed in.}
  \label{fig:overview}
\end{figure}

\section{An overview}
\label{sec:overview}
In the previous sections, we have described the individual parts of \libsme{}
without describing the integration of its components. Hence, we devote a section
for that purpose here. An overview of the \libsme{} library and its interactions
is shown in \Cref{fig:overview} and the individual stages are described below.

\subsection{Parsing and Import Resolution} Regardless of how \libsme{} is
invoked (as described in the previous section) the SMEIL source is parsed and
the resulting AST is passed through the import resolver. Here, the code is
scanned for the presence of {\ttfamily import} statements. If any are found, the source
files containing the imported modules are parsed in a recursive manner. Whenever
we recursively import a module, we pass along the list of imported entities such
that only the requested entities are imported. The tree of imported modules is
then flattened by renaming hierarchical references. This process seeks to
simplify the subsequent phases of the compilation process as module hierarchies
do not have to be considered. The renamings are tracked and passed on to the
following stages so that a reverse mapping may be performed later, for example
for error messages.

% In more detailed terms, the algorithm for performing imports work as
% follows:\todo{Merge this paragraph with the previous} A
% module import is handled through recursive descendant evaluation of imports
% where different information is passed on the outgoing and incoming edges of the
% module dependency graph. On forward edges, we pass the import
% ``paramteres'' of a module and on the backward edges we pass the renamed
% module. Every time something is passed on an incoming edge, it is folded (merge)
% into the code on that level. Before merging, all top-level names and references
% to those names are renamed such that there is no name-clashes with the code that
% the module is being merged into. During this process, we also ensure that all
% imported names actually exists and produce a error if they do not.

% The process of importing a module goes through several steps. The aim of this
% process is to end up with a flattened and normalized module and thus to discard
% the module information as early as possible in the compilation
% process. Conceptually, the process mimics the recursive nature of the module
% system where different information flows forward and backwards along the forward
% and backwards edges of the recursion call-tree. Module system implementations
% more commonly aim to treat modules as isolated units


\subsection{Type Checking} The code is then passed through the type checker
which enforce the typing rules described in \Cref{sec:typesys}. The type checker
makes two passes through the code:
\begin{itemize}
\item The first pass locates all entity definitions (processes and networks) and
  adds them to the top-level symbol table. For every entity found, the
  declarations in that entity are added to a local symbol table which is
  associated with the entity.
\item The second pass performs type checking on all declarations and statements
  in the previously discovered entities. During this process, the individual
  AST-nodes are annotated with their types.
  %\todo{Should there be an example here?}
  Having such type information available throughout the AST is of significant
  value for subsequent passes, such as code generation and simulation, since
  they are able to determine the type of an expression at any time by looking at
  its AST node.
%     For example, suppose
%     expression {\ttfamily 2 + 6} is represented in an AST as
% \begin{verbatim}
% (PlusOp Untyped (LitInt 2) (LitInt 6)

% \end{verbatim}
\end{itemize}

The two-pass approach ensures that declarations can be given in any
order. Requiring declarations to be made ahead-of-use would make the code shown
in many of the examples shown throughout this thesis become significantly more
convoluted.

A single abstract representation of SMEIL is used throughout the compiler. Code
simplifications could be made if an intermediate representation of SMEIL was
used by the internal stages of the compiler. However, introducing such an
intermediate representation would limit our ability to reconstruct the original
SMEIL source code following re-typing (\Cref{sec:typing}). Furthermore,
maintaining an unchanged representation of the original source code means that
the generated code more closely corresponds to the source code.

\subsection{Analysis} The analysis phase examines the structure of a network.
This is used for determining the top-level entity of the network which is needed
both for deriving a runtime representation of the network and for subsequent
code generation. From here, the AST may take two paths depending on the mode of
invocation requested by the user. It is either passed on directly to the code
generator, or simulated. If the AST was already retyped by the simulator, it is
passed directly to the code generator.

\subsection{Simulation}
Simulation is performed to test a design. During the simulation, the value
ranges assigned to every variable and bus channel are tracked such that we can
use them for constraining integer types. Furthermore, the values of
external-facing buses are logged and used to construct the CSV trace file used
by the generated VHDL test bench.
%\todo{Discuss why we only have external interfaces}
The simulator also performs accurate emulation of integer overflows. During
simulation, if \libsme{} is used for co-simulation with another SME network, it
will exchange the values of external-facing buses with another SME
network. After simulation, the AST may either be passed directly to the code
generator or, if new types were assigned, returned to the type checker.

In very early phases of this project, we considered if implementing a simulator
for SMEIL was even needed. After all, if SMEIL is used as an intermediate
language, the source SME network could be simulated directly leaving SMEIL to be
used purely for code generation. In this scenario, the trace file used for the
test bench would simply be passed along with the SMEIL intermediate code and
used for providing input to the generated VHDL test bench. However, we
determined that without a simulator, SMEIL would be restricted to this
particular use case only.

\subsection{Code generation}
\label{sec:codegen}
The final stage, yielding the desired output, is the code generation phase
which, as its name suggests, turns the typed and possibly simulated SMEIL AST
into VHDL code.

SMEIL compiles to clean and readable VHDL code which is amenable to manual
modifications. The code may be executed using a VHDL simulator or passed to FPGA
vendor tools for synthesis and subsequent hardware-implementation (as described
in \Cref{sec:sme}). The generated code is a cycle-accurate representation of the
original SMEIL network.

The fundamental structure of the SMEIL code is preserved in the generated VHDL
code. One VHDL entity is generated per SMEIL entity and the body of an SME
process is transformed into a VHDL {\tt architecture} containing a single
sequential process. For each of these processes, we also generate code for
performing an asynchronous reset of all variables and outgoing
signals. The naming hierarchy of the original SMEIL is preserved, making it easy
to identify from where a particular section of the VHDL code was generated. In
\Cref{sec:colorbin} we show an example of how a SMEIL process is transformed
into an FPGA entity.

For verifying the generated VHDL code, a test bench is also generated. The
test bench is a VHDL program which connects to the {\ttfamily exposed} buses of
the SMEIL program. The CSV-trace file, containing the values recorded during
simulation, is used by the VHDL test bench to drive inputs and verify outputs.

Alongside the generated code, a {\ttfamily Makefile} is generated for building
and testing the VHDL code using the GHDL~\cite{ghdl} simulator.

Integer types of SMEIL are represented in VHDL using the types provided by the
standard {\ttfamily ieee.numeric\_std} package. This package provides functions
for performing signed and unsigned integer arithmetic with logic-vectors. For
example, the types {\ttfamily i4} and {\ttfamily u12} are represented as
{\ttfamily signed (3 downto 0)} and {\ttfamily unsigned (11 downto 0)}
respectively.
% Contrary to SMEIL, t, that it can not overflow. As we
% described in \Cref{sec:typesys} this often leads to an over-estimation of bits
% required to hold a value.

Arrays require the creation of a new {\ttfamily type} in VHDL. A {\ttfamily
  type} declaring a 10-element array of 5-bit signed integers ({\ttfamily
  [10]i5} in SMEIL) is represented in VHDL as
\begin{lstlisting}[language=vhdl]
type \[10]i5\ is array (0 to 9) of unsigned (4 downto 0);
\end{lstlisting}
These type declarations are stored in a separate package, {\ttfamily sme\_types.vhdl},
which is shared between all entities of the design to avoid cluttering the
generated code with duplicated declarations. SMEIL booleans are represented
using the VHDL type {\ttfamily boolean}. As an alternative to this, a single {\ttfamily
  std\_logic} type is commonly used. This type represents a wire in the hardware
and is, therefore, able to have other states than just true or false. This may
be useful in some circumstances.

The actual code is generated using the {\ttfamily language-vhdl-quote}
library~\cite{language-vhdl-quote} which provides
quasiquoters~\cite{mainland2007s} for building VHDL ASTs using the concrete VHDL
syntax. The major advantage of this approach is that it minimizes the chance of
generating syntactically invalid VHDL since syntax errors are caught during
compilation of \libsme{}. Furthermore, a complete VHDL AST is constructed
containing the contents of each generated VHDL file. This AST is then
pretty-printed, yielding consistently formatted code which is difficult to
achieve using more common techniques based on string templates.

\subsection{Reconstruction} If observation based typing was enabled, the
simulator will have annotated the SMEIL AST with types based on the observed
values. By reconstructing a structure resembling the original AST, reusing the
stages of the compiler is simplified. Furthermore, the results of the retyping
are shown to the user using nicely formatted concrete SMEIL syntax.

% \section{Runtime Representation of SMEIL}
% Since entities in SMEIL can be instantiated, there is not a one-to-one
% correspondence between the number of entity and bus declarations and the number
% of objects in the runtime representation. For example, the following network
% \begin{smeilcode2}
% proc A ()
%     instance b1 of B();
%     instance b2 of B();
% {}

% proc B ()
%     instance c1 of C();
%     instance c2 of C();
%     instance c3 of C();
% {}

% proc C ()
% {}

% network N () {
%     instance _ of A();
% }
% \end{smeilcode2}
% forming a tree of instances, will at runtime contain 6 instances of the {\tt C}
% process even though only 3 {\tt instance} declaration for instantiating it are
% present in the code. Therefore, the simple solution of scanning the code for
% {\tt instance} statements and creating one process instance per occurrence will
% not work.

% % Here, we describe the algorithm used
% % to go from program declaration to runtime objects. \todo{finish}

% % Like the handling of import statements, the algorithm for establishing a runtime
% % representation of SMEIL 

% % \todo{Describe the recursive algorithm used for creating a runtime
% %   representation of an SMEIL}

% % The key insight of this algorithm is that a network, connecting two processes,
% % for example

% Another problem to handle is seen in the following network block
% \begin{smeilcode}
%   network n() {
%     instance i1 of p1(p2.b);
%     instance i2 of p2(p1.b);
%   }
% \end{smeilcode}
% which instantiates two processes i1 and i2 and creates a connection between
% them. What makes this hard, is that even though the connection between the two
% processes is established in the network {\tt n}, the buses which forms the
% connections are defined within {\tt p1} and {\tt p2} processes. Therefore, we
% need to create the {\tt i1} and {\tt i2} instances in order to know which buses
% to connect.

% % Another thing to note is that networks have no effect on the runtime
% % representation of an SMEIL program.

% The input to the process is an SMEIL program and the result is a graph
% containing process instances with links to the buses that connects
% them.


% This runtime representation is established through the following
% algorithm:
% \begin{enumerate}
% \item For every {\tt instance} declaration encountered referencing a process we
%   create a runtime process containing a symbol table. We fill this symbol table
%   with all declarations {\itshape except} instance found in the declarational
%   part of the process.
    
%   \item For every such newly created tree node we create a bus instance for
%     every bus declaration in the entity.
%   \item Then, every instance
%   \item When
% \end{enumerate}

% For example, in the code shown above, we would instantiate the processes

% \section{Design challenges}
% The key challenge faced in the design of libsme was that, at any stage in the
% compiler, we had to be able to return to the original representation.
% TODO: Discuss issues related to maintaining a single internal representation
% throughout the compiler which is reconstructable to the original concrete
% syntax.

% Early on, we considered a particular challenge that this project would
% pose. Primarily due to the following design goals

% An ongoing dilemma throughout the design of \libsme{} were the balance of code
% complexity versus usability. As seen in the overview of the SMEIL design
% \todo{reference} there are numerous optional syntactical features. An idea that
% we have been conti


% \todo{evaluate impl. quality}
% The . Another area where statically typed functional languages excel is code
% refactoring. For a large program written in, for example, Python, changing a
% fundamental data structure is a daunting task. In Haskell, you simply change the
% data structure to the desired form and fix the resulting type errors throughout
% your program.

% In contrast with other, impure, functional languages, such as F\# or ocaml, the
% purity of Haskell means that mutable state is introduced in a highly controlled
% manner, using monadic compositions. This is in contrast to dynamic languages
% such as Python where 100\% test coverage is needed to ensure that a typo in a
% rarely traversed code path will not cause the program to crash. Overall, our
% experience developing a moderately large project in Haskell has been extremely
% pleasant. By using Haskell, we have traded a slightly slower development pace,
% for significantly less time spent debugging.

% \todo{section}
\section{Software-engineering considerations}
The language chosen for implementation of \libsme{} is Haskell. It would have
been possible to carry out the implementation in any general-purpose language,
but Haskell was chosen in particular because:
\begin{itemize}
\item Functional programming languages are well suited for writing
  compiler-related software, due to their support for Algebraic Data Types
  (ADTs) and pattern-matching. Also, a wide range of libraries exists for
  supporting the implementation of for example parsers and pretty-printers.
\item The type-safety of Haskell trades a slightly slower development pace for a
  significant reduction of time spent debugging.
\item The type system also significantly aids refactoring, something which
  proved useful several times while developing this project. When a data
  structure is changed in a Haskell program, the type system ensures that
  compile-time errors are raised for code affected by the change.
  % For
  % example, changing a fundamental data structure in Python can be a daunting
\end{itemize}

\libsme{} comprises just short of 6000 SLOC of Haskell. Additionally, the
wrapper module for holding the co-simulation state and neatly exposing the
functions of the C-API is implemented in a module is approximately 500 SLOC of
C. The VHDL parsing and quasiquotation library developed for use with this
project consists of approximately 5500 SLOC of Haskell.

The implementation currently has several rough edges, but its fundamental
structure is sane and it has been written with future extensions in mind. It
also pays particular attention to usability-related features such as providing
understandable error messages.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../master"
%%% End:
